Neural Connectome of the Ctenophore Statocyst

### 2025-06-25 manuscript/Manuscript.qmd
MS: reference all supllementary figures and videos in main text

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 898 of which formed the aboral organ (AO). Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

@@

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Figure 2--figure supplement 1C). ANN neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023]. In our dataset, we also reconstructed 10 SNN neurons. These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1). Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement C-G).

@@

Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant. These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth. The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 mitochondria.

@@

The second group of cells that formed synaptic contacts with the ANN were the bridge cells. Bridge cells, first described in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells. Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells (Figure 3 -- figure supplement 1).

@@

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figures supplement 1). These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.

@@

We found that arrest and re-beat events were synchronized between balancers across the sagittal plane. However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.16 seconds (Figure 4D). Overall, these data reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4, whereas re-beat is coordinated over all four quadrants.

As shown in Figure 4E, beat frequencies of left and right balancers fluctuated substantially in both orientations. However, while left-right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) showed strikingly periodic alternations between left and right balancer activity. This pattern suggested an offset in timing during arrest or re-beat events between tentacular balancers.

@@

![**Figure 4—figure supplement 1.Ciliary beat frequency (CBF) dynamics of balancer cells in additional samples**\
Bar plots showing CBF dynamics of left and right balancers (top and bottom panels, respectively) recorded from the sagittal (A–C) and tentacular (D–F) planes. Positive and negative values indicate opposing sides within each imaging plane. These data correspond to additional samples not shown in the main figure and support the results presented in Figure 4E.](figures/Figure4_Supplement1.png)

>>>>

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 898 of which formed the aboral organ (AO). Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ. We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

@@

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Figure 2--figure supplement 1C). ANN neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023]. In our dataset, we also reconstructed 10 SNN neurons. These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1). Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement 1 C-G).

@@

Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant (Video 2). These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth. The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 mitochondria.

@@

The second group of cells that formed synaptic contacts with the ANN were the bridge cells. Bridge cells, first described in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells. Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells (Figure 3 -- figure supplement 1, Video 1).

@@

The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, epithelial floor cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figure supplement 1). These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.

@@

We found that arrest and re-beat events were synchronized between balancers across the sagittal plane. However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.16 seconds (Figure 4D, Videos 3-6). Overall, these data reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4, whereas re-beat is coordinated over all four quadrants.

As shown in Figure 4E (and Figure 4--figure supplement 1), beat frequencies of left and right balancers fluctuated substantially in both orientations. However, while left-right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) showed strikingly periodic alternations between left and right balancer activity. This pattern suggested an offset in timing during arrest or re-beat events between tentacular balancers.

@@

![**Figure 4—figure supplement 1. Ciliary beat frequency (CBF) dynamics of balancer cells in additional samples**\
Bar plots showing CBF dynamics of left and right balancers (top and bottom panels, respectively) recorded from the sagittal (A–C) and tentacular (D–F) planes. Positive and negative values indicate opposing sides within each imaging plane. These data correspond to additional samples not shown in the main figure and support the results presented in Figure 4E.](figures/Figure4_Supplement1.png)
